---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bnpe

<!-- badges: start -->
`r badger::badge_license(color = "orange")`
`r badger::badge_lifecycle("experimental", color = "blue")`
<!-- badges: end -->

'bnpe' is an R-package proposes a collection of function to help retrieve the French national data bank of quantitative withdrawals (Banque Nationale des Prélèvements quantitatifs en Eau - BNPE) available on its website: https://bnpe.eaufrance.fr

## Installation

You can install the bnpe R-package from it's development repository with:

``` r
install.packages("remotes")
remotes::install_gitlab("in-wop/bnpe", host = "gitlab.irstea.fr")
```

## Get started

### Loading library

```{r library}
library(bnpe)
```

### Time series

Functions for getting time series are: 

- `getTimeSeriesDep`: for one department
- `getTimeSeriesCom`: for one commune

These functions uses the French official geographical codification (https://fr.wikipedia.org/wiki/Code_officiel_g%C3%A9ographique).

Type `?getTimeSeriesCom` for getting help.

Examples:

```{r timeseries}
# Time series for all type of withdrawal sources and destinations in the Ardennes departement
getTimeSeriesDep("08")

# Time series for drinking water withdrawals in surface water bodies in the Ardennes department
getTimeSeriesDep("08", code_usage = "AEP", code_type_eau = "CONT")

# Withdrawal time series for the commune of Charleville-Mézières
getTimeSeriesCom("08105")
```

###  Communal data in one departement

For getting list of communes of departement with corresponding withdrawal volumes for one year, use the function `getComSeriesDep` as follow:

```{r}
# All source and destination withdrawals for the year 2016 in the Ardennes departement
getComSeriesDep("08", 2016, code_usage = "AEP", code_type_eau = "CONT")
```

### Properties and time series from one device ("ouvrage")

For getting metadata and withdrawal yearly time series from on device identified by its "code Sandre" (See https://www.sandre.eaufrance.fr/atlas/srv/fre/catalog.search#/metadata/e315633f-0930-41e8-a1c4-61fb2303039c):

```{r}
# metadata and timeseries from withdrawal "COPRIMANCHE-forage LD LE PACO (40m)" 
ouvrage <- getOuvrageSeries("OPR0000200109")
str(ouvrage)
```
