---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r opts, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# hubeau an R package for the Hub'eau APIs

<!-- badges: start -->
`r badger::badge_license(color = "orange")`
`r badger::badge_lifecycle("experimental", color = "blue")`
<!-- badges: end -->

'hubeau' is an R-package proposing a collection of function to help retrieve data from the French national databases on water: https://hubeau.eaufrance.fr

## Installation

You can install the 'hubeau' R-package from it's development repository with:

``` r
install.packages("remotes")
remotes::install_gitlab("in-wop/hubeau", host = "gitlab.irstea.fr")
```

## Loading library

```{r library}
library(hubeau)
```

## Get started

See the package documentation by typing:


```{r}
?hubeau
```

Each function is documented with at least one example: 

```{r}
example("get_prelevements_chroniques")
```


