---
title: "Querying the API 'Ecoulement'"
author: "Pascal Irz, David Dorchies"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ecoulement_cours_deau}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning   = FALSE,
  message   = FALSE
)
```


# Aim

This vignette shows how to use the [`hubeau` R package](https://github.com/inrae/hubeau) in order to retrieve data from the French streams drought monitoring network ([Observatoire National Des Etiage, ONDE](https://onde.eaufrance.fr/)) from the [API "Ecoulement des cours d'eau"](https://hubeau.eaufrance.fr/page/api-ecoulement) of the [Hub'eau](https://hubeau.eaufrance.fr/) portal. This is illustrated for the departement of Ille-et-Vilaine (Id: 35).

Nationwide, this network encompasses 3000+ observation sites, that are streams stretches selected to reflect the severity of drought events, i.e. that flow in "normal" conditions but dry out in case of drought. Being upstream in the river basins, these sites are expected to act as 'sentinels' and to provide early warning of the necessity of restricting water withdrawal in rivers and associated water tables. 

```{r}
my_dept <- "35"
```

# Setup

Install package `hubeau`.

```{r, eval = FALSE}
# Install released version from CRAN
install.packages("hubeau")

# Install development version from GitHub:
# install.packages("remotes")
remotes::install_github("inrae/hubeau")
```

Load packages.

```{r setup}
library(hubeau)
library(dplyr)
library(sf)
library(mapview)
library(ggplot2)
library(forcats)
```


# Getting started

"Ecoulement" is one of the `r length(hubeau::list_apis())` APIs that can be queried with the `{hubeau}` R package. These APIs can be listed using the `list_apis()` function.

```{r}
hubeau::list_apis()
```

To know the possible endpoints of the API, use the `list_endpoints` function.

```{r}
hubeau::list_endpoints(api = "ecoulement")
```

- `stations` lists the monitoring stations
- `campagnes` lists the surveys
- `observations` is the data itself, indicating if, at the date of the survey, the river flows or if it is dry (which is assessed visually in the field).

`station` and `observations` can be joined *at least* by the field `code_station`.

`observations` and `campagne` can be joined by the field `code_campagne`.

For each endpoint, the function `list_params()` gives the different parameters that can be retrieved. 

```{r}
hubeau::list_params(api = "ecoulement",
                    endpoint = "observations")
```

# Retrieving the data

## Sites

Retrieve the available fields.

```{r}
param_stations <- 
  list_params(api = "ecoulement",
              endpoint = "stations") %>% 
  # needs a bit of cleaning
  toString() %>% 
  gsub(pattern = " ",
       replacement = "")
```

Download the data.

```{r}
stations <-
  get_ecoulement_stations(
    code_departement = my_dept,
    fields = param_stations)
```

The `stations` dataframe is transformed into a `sf` geographical object, then mapped. 

```{r, fig.height = 5, fig.width = 6}
stations_geo <- stations %>% 
  select(code_station,
         libelle_station,
         longitude,
         latitude) %>% 
  sf::st_as_sf(coords = c("longitude", "latitude"),
               crs = 4326)

mapview::mapview(stations_geo,
                 popup = leafpop::popupTable(stations_geo,
                                    zcol = c("code_station", "libelle_station"),
                                    feature.id = FALSE,
                                    row.numbers = FALSE))
```


## Surveys

A survey encompasses a series of visual observations carried out at on stations at the "departement" grain. It is tagged as "usuelle" if it is a routine (i.e. around the 24th of the month from May to September), as "complémentaire" if not (e.g. during very dry periods).

Download the `surveys` dataframe.

```{r}
surveys <- get_ecoulement_campagnes(
  code_departement = my_dept, # department id
  date_campagne_min = "2012-01-01" # start date
  ) %>%
  mutate(code_campagne = as.character(code_campagne)) %>%
    select(code_campagne,
           date_campagne,
           libelle_type_campagne)
```

Visualise a few rows.

```{r}
surveys %>%
  head() %>% 
  knitr::kable()
```

## Observations

Retrieve the available fields.

```{r}
param_obs <- 
  list_params(api = "ecoulement",
              endpoint = "observations") %>% 
  toString() %>% 
  gsub(pattern = " ",
       replacement = "")
```

Download the data.

```{r}
observations <- 
  get_ecoulement_observations(
    code_departement = my_dept,
    date_observation_min = "2012-01-01",
    fields = param_obs
                       ) %>%
  mutate(code_campagne = as.character(code_campagne))
```

The flow observations are codes into discrete levels:

```{r}
observation_levels <- observations %>%
  select(code_ecoulement,
         libelle_ecoulement) %>%
  distinct() %>%
  arrange(code_ecoulement)

english_labs <- data.frame(
  ENG_lab = c(
    "Visible flow",
    "Decent visible flow",
    "Weak visible flow",
    "No visible flow",
    "Dry"
  )
)

englishisized_labs <- observation_levels %>%
  bind_cols(english_labs)

flextable::flextable(englishisized_labs)
```

# Assembly in a single dataframe


```{r}
data <- observations %>% 
  left_join(surveys) %>%
  left_join(stations) %>%
  left_join(y = englishisized_labs) %>% 
  mutate(date_campagne = as.Date(date_campagne),
         year = lubridate::year(date_campagne),
         month = lubridate::month(date_campagne),
         ENG_lab = as.factor(ENG_lab),
         ENG_lab = fct_relevel(ENG_lab,
                                "Visible flow",
                                "Decent visible flow",
                                "Weak visible flow",
                                "No visible flow",
                                "Dry",
                                NA))

data2 <- data %>% 
  select(code_station, year, month, libelle_station, ENG_lab) %>% 
  tidyr::complete(year, month, fill = list(ENG_lab = NA))
```

# Graphical output

## For a station

Custom plot function.

```{r}
gg_stream_flow <-
  function(data, sel_station) {
    # selected data
    sel_data <- data %>%
      filter(code_station == sel_station)
    
    # station name for plot title
    station_lab <- unique(sel_data$libelle_station)
    
    # year range
    year_range <-
      min(sel_data$year, na.rm = T):max(sel_data$year, na.rm = T)
    
    # complete df with the months without data
    dates <- data.frame(year = rep(year_range, 12) %>% sort(),
                        month = rep(1:12, length(year_range)))
    
    sel_data <- dates %>%
      left_join(sel_data)
    
    # color palette
    colors <- c(
      "Visible flow" = "blue",
      "Weak visible flow" = "lightblue",
      "No visible flow" = "orange",
      "Dry" = "red"
    )
    
    # plot
    sel_data %>%
      ggplot(aes(x = year,
                 y = month,
                 color = ENG_lab)) +
      geom_point(shape = 15, size = 7) +
      coord_flip() +
      scale_color_manual(name = "",
                         values = colors,
                         na.value = "grey80") +
      scale_y_continuous(breaks = 1:12,
                         labels = 1:12,
                         limits = c(1, 12)) +
      scale_x_continuous(breaks = year_range,
                         labels = year_range) +
      labs(x = "",
           y = "Month",
           title = station_lab) +
      theme(
        axis.line = element_line(color = 'black'),
        plot.background = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_line(linewidth = 0.5,
                                          colour = "black"),
        panel.border = element_blank()
      )
    
  }

```

Display plot for a station.

```{r, fig.width = 4.5, fig.height = 4}
my_station <- "J7364221"

gg_stream_flow(data = data2,
               sel_station = my_station)
```
