---
title: "Querying the API 'Ecoulement'"
author: "Pascal Irz, David Dorchies"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ecoulement_cours_deau}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning   = FALSE,
  message   = FALSE
)
```

```{r setup}
library(hubeau)
library(tidyverse)
library(sf)
library(mapview)
# library(leaflet)
# library(leaflet.extras)
# library(ggrepel)
# library(glue)
# library(forcats)
# library(scales)
# library(data.table)
```

# Aim

This vignette shows how to retrieve data from the French drought monitoring network ([Observatoire National Des Etiage, ONDE](https://onde.eaufrance.fr/)) from the [API "Ecoulement des cours d'eau"](https://hubeau.eaufrance.fr/page/api-ecoulement) of the [Hub'eau](https://hubeau.eaufrance.fr/) portal. The exemple is given for the departement of Ille-et-Vilaine (Id: 35).

```{r}
my_dept <- "35"
```


# Getting started

"Ecoulement" is one of the `r length(hubeau::list_apis())` APIs that can be queried with the `{hubeau}` R package, that can be listed using the `list_apis()` function.

```{r}
hubeau::list_apis()
```

To know the possible endpoints of the API, use the `list_endpoints` function.

```{r}
hubeau::list_endpoints(api = "ecoulement")
```

- *stations* lists the monitoring stations
- *campagnes* lists the surveys
- *observations* is the data itself, indicating if, at the date of the survey, the river flows or if it is dry (which is assessed visually in the field).

These 3 elements can be joined 

For each endpoint, the function `list_params()` gives the different parameters that can be retrieved. 

```{r}
hubeau::list_params(api = "ecoulement",
                    endpoint = "observations")
```

# Retrieving the data

## Sites

```{r}
param_stations <- 
  list_params(api = "ecoulement",
              endpoint = "stations") %>% 
  toString() %>% 
  gsub(pattern = " ",replacement = "")

stations <-
  get_ecoulement_stations(
    code_departement = my_dept,
    fields = param_stations)

stations_geo <- sf::st_as_sf(x = stations,
                             coords = c("longitude", "latitude"),
                             crs = 4326)

mapview::mapview(stations_geo)
```


## Surveys

A survey encompasses a series of visual observations carried out at on stations at the "departement" grain. It is tagged as "usuelle" if it is a routine (i.e. around the 24th of the month from May to September), as "complémentaire" if not (e.g. during very dry periods).  

```{r}
surveys <- get_ecoulement_campagnes(
  code_departement = my_dept, # department id
  date_campagne_min = "2012-01-01" # start date
  ) %>%
  mutate(code_campagne = as.character(code_campagne)) %>%
    select(code_campagne,
           date_campagne,
           libelle_type_campagne)
  
surveys %>%
  head() %>% 
  knitr::kable()
```





```{r}
#### Observations
param_obs <- 
  list_params(api = "ecoulement", endpoint = "observations") %>% 
  toString() %>% 
  gsub(pattern = " ",replacement = "")

observations <- get_ecoulement_observations(
                         code_departement = my_dept,
                              date_observation_min = "2012-01-01",
                              fields = param_obs
                       ) %>%
  mutate(code_campagne = as.character(code_campagne))
```


```{r}
### Assembly in a single dataframe
onde_df <- observations %>% 
  left_join(surveys) %>%
  left_join(stations) %>%
  mutate(date_campagne = as.Date(date_campagne),
         Annee = lubridate::year(date_campagne)) %>%
  arrange(code_station,code_departement,desc(Annee))
                

```

