---
title: "Performing queries on the 'niveaux_nappes' API"
author: "Pascal Irz & David Dorchies"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example_niveaux_nappes_api}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "70%",
  fig.align = "center"
)
load(system.file("vignettes/example_niveaux_nappes_api.RData", package = "hubeau"))
```

This vignette shows how to use the [`hubeau` R package](https://github.com/inrae/hubeau) in order to retrieve data from the French piezometric monitoring network ([portail national d'Accès aux Données sur les Eaux Souterraines, ADES](https://ades.eaufrance.fr/)) from the [API "piézométrie" (or "niveaux nappes")](https://hubeau.eaufrance.fr/page/api-piezometrie) of the [Hub'eau](https://hubeau.eaufrance.fr/) portal. 

We illustrate the use of this API for the geological layer "Calcaires et marnes du Dogger du bassin versant du Clain libres" (`code=GG063`) with an example of map and chart useful for the interpretation of these data.

```{r}
my_water_table_code <- "GG063"
```

# Getting started

First, we need to load the packages used in this vignette for processing data and
display results on charts and map:

```{r setup}
library(hubeau)

library(dplyr)
library(sf)
library(mapview)
library(ggplot2)
library(purrr)
```

"niveaux_nappes" is one of the `r length(hubeau::list_apis())` APIs that can be queried with the `{hubeau}` R package.

The list of the API endpoints is provided by the function `list_endpoints`.

```{r}
list_endpoints(api = "niveaux_nappes")
```

- `chroniques` long-term time series
- `chroniques_tr` real time observations
- `stations` lists the monitoring stations

The tables can be joined *at least* by the field `code_bss` (i.e. the site Id).

For each endpoint, the function `list_params()` gives the different parameters that can be retrieved. 

```{r}
list_params(api = "niveaux_nappes",
            endpoint = "stations")
```

# Retrieving the data

## Sites

Download the data.

```{r, eval = FALSE}
stations <- get_niveaux_nappes_stations(
  codes_masse_eau_edl = my_water_table_code
)
```

## Time series

The aim here is to retrieve the time series of water levels measures in the piezometers.

```{r}
param_chroniques <- paste(
    list_params(api = "niveaux_nappes",
                endpoint = "chroniques"),
    collapse = ","
)
```

No field allows to select a departement, so it is necessary to iterate through stations. 

```{r, eval = FALSE}
water_table_level <- map_df(
  .x = stations$code_bss,
  .f = function(x)
    get_niveaux_nappes_chroniques(code_bss = x,
                                  date_debut_mesure = "2015-01-01")
)
```

# Tidying the data

The dataframe is processed to get a 'year' and a 'month' variables, then averaged by year for each site.

```{r, eval = FALSE}
water_table_level <- water_table_level %>% 
  mutate(date_mesure = lubridate::ymd(date_mesure),
         year = lubridate::year(date_mesure),
         month = lubridate::month(date_mesure))
```

Selection, for each of the stations, of the years with 12 months of data. 
This is done to prevent incomplete time periods to influence the yearly mean water level excessively.

```{r, eval = FALSE}
yearly_mean_water_table_level <- water_table_level %>% 
  group_by(code_bss,
           year) %>% 
    summarise(n_months = n_distinct(month)) %>% 
    filter(n_months == 12) # complete years

yearly_mean_water_table_level <- yearly_mean_water_table_level %>% 
  select(-n_months) %>% 
  left_join(water_table_level) %>% # filtering join
  group_by(code_bss,
           year,
           month) %>% 
    summarise(monthly_mean_water_table_level = mean(niveau_nappe_eau, na.rm = TRUE)) %>% 
  group_by(code_bss,
           year) %>% 
    summarise(yearly_mean_water_table_level = mean(monthly_mean_water_table_level, na.rm = TRUE)) %>% 
  ungroup()
```

# Plotting

```{r, fig.width = 8, fig.height = 8}
ggplot(data = yearly_mean_water_table_level,
       aes(x = year,
           y = yearly_mean_water_table_level)) +
  geom_line() +
  facet_wrap(~code_bss,
             scales = "free_y")
         
```

# Mapping

The `stations` data.frame is transformed into a `sf` geographical object.

```{r}
stations_geo <- stations %>% 
  st_as_sf(coords = c("x", "y"),
           crs = 4626)
```

We create a plot for each station ready to be displayed as map "pop-up".

```{r}
p <- lapply(unique(yearly_mean_water_table_level$code_bss),
            function(x) {
              ggplot(data = yearly_mean_water_table_level %>% filter(code_bss == x),
                     aes(x = year,
                         y = yearly_mean_water_table_level)) +
                geom_line() +
                labs(title = x)
            })
```

Then they are mapped using the `mapview` R package. Click on a spot to popup the
plot.

```{r, out.width = "100%", fig.asp = 1}
mapview(
  stations_geo,
  map.types = c("OpenStreetMap",
                "Esri.WorldShadedRelief",
                "OpenTopoMap"),
  popup = leafpop::popupGraph(p)
)
```
